`respR` - An R package for the analysis of respirometry data

### 2017-10-17 manuscript/manuscript-HTML.Rmd
manuscript update
tried to fix equations but not sure if they are right, yet

Metabolic rate is a fundamental trait linked to virtually all other biological functions, and is key to the prediction of patterns in ecology and conservation biology, from populations (Seibel and Drazen, 2007, Barneche et al., 2014) to ecosystems (Brown et al., 2004). It has been widely used to investigate the effects of external stressors on an organism, and has seen increased popularity in studies involving climate change where warming is expected to drive increased metabolism in ectotherms (Pörtner, 2002, McElroy et al., 2012, Carey et al., 2016, Delorme & Sewell, 2016). By far the most common metric used to measure physiological performance in animals, it has been applied from the whole organism to the level of cells and tissues (White & Kearney, 2013), and is typically quantified using respirometry, which measures oxygen uptake over time (Lighton, 2008).

There are at least four broad methodological approaches in respirometry: closed, intermittent flow, flowthrough and open. In closed respirometry, $O_2$ decrease is measured within a hermetically sealed chamber of known volume, sometimes set within a closed loop to allow circulation or mixing of the environment within the chamber. The oxygen recordings may be continuous through use of an oxygen probe, periodic through withdrawing water or gas samples at set intervals, or a two-point measurement consisting of the initial and final concentrations. Metabolic rates are estimated from respirometry by assuming a linear relationship between variables, and estimates of metabolic rate are straightforward in constant volume respirometry using the equation: $$VO_2 = \Delta O_2(V)$$ where $\Delta O_2$ is the slope of the regression that relates O~2~ concentration to time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time, and $V$ is the volume of the container (Lighton, 2008).

@@

Flowthrough respirometry involves a closed chamber, but with a regulated flow of air or water through it at a precisely determined rate. After equilibrium has been achieved, the oxygen concentration differential between the incurrent and excurrent channels, along with the flow rate, allows calculation of the oxygen extracted from the flow volume per unit time: $$VO_2 = (C_iO_2 - C_eO_2)FR$$ where $VO_2$ is the rate of $O_2$ consumption in milligrams per minute, $C_iO_2$ and $C_eO_2$ are the incurrent and excurrent $O_2$ concentrations, and $FR$ is the flow rate of water through the system  (Lighton, 2008).

A final method is open respirometry, in which an open tank or semi-enclosed area is used, but the input or mixing rate of oxygen from the surroundings is known or found to be negligible relative to oxygen consumption of the specimens (Leclercq et al., 1999). It is seldom used, but for some applications it is a sufficient and practical methodology (e.g. Gamble et al., 2014). The common equation used for open respirometry is: $$VO_2 = \Delta O_2 + \phi_d$$ where $\Delta O_2$ is the slope of the regression that relates $O_2$ concentration to time, and $\phi_d$ is the air-sea oxygen flux as determined by Fick's Law (Leclercq et al., 1999).

@@

Data should be formatted correctly before use in `respR`. The function `check.input()` performs error checks on a data frame of any size, and can extract a two-column data frame for data exploration and subsequent analyses. Data must be numeric, because `respR` has limited support for date/time data (e.g. `POSIXct` and `POSIXlt` classes) and performing calculations on different time formats have many caveats that are beyond the scope of the package. Time data should be sequential, without duplicates, as that might signify instrumental drift. Ideally, time data should also be evenly-spaced, that is, the time period between samples are equal. This has important ramifications on data extraction techniques later on as substantial errors may occur in the event that calculations are based on the number of rows instead of time. A plot of the data is also shown for visual inspection (e.g. Fig. 1).

@@

Background $RO_2$ is measured to account for the influence of bacterial respiration in the chamber, and once measured, is often applied as a correction to the main experiment (Rodgers et al., 2016). Since background rates typically account for a small percentage of experimental rates, these “blank” experiments are routinely conducted alongside main experiments, and the rates are averaged across several datasets to obtain a more accurate estimate of the correction. The function `calc.bg.rate()` can simultaneously process multiple background rate measurements, as long as they share the time data, and produces an output object that can be used in later functions that accept background rate arguments.

@@

A typical procedure in the analysis respirometry data is to select the most linear section of the data for calculations of $RO_2$ (e.g. citations). The function `calc.rate()` can manually extract one or more segments from a single data frame to process them simultaneously. Data segments can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a regular 10-minute time window across multiple specimens).

In addition, we can take advantage of automated methods in `auto.rate()`, which implements novel rolling regression and kernel density estimation algorithms to detect patterns in the data to automatically obtain the maximum or minimum rate of fixed width, or to detect the most linear sections of the data. Unlike manual methods where data is most likely arbitrary selected, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the entire data frame (Fig. 2b). Thus, a total of $(m−n)+1$ number of iterative regressions are fitted, and we obtain a new dataset of regressions as a function of time (Fig. 3), which is then ranked by size to obtain maximum and minimum values.

To estimate the most linear rate, further computations are performed after rolling regressions. Logically, linear sections of the data should reflect a consistently stable (i.e. flat) value in the rolling regression data (Fig. 3b). The function `auto.rate()` takes advantage of this relationship to identify data windows that contain consistent rates over a rolling window, implementing the methods of Sheather & Jones (1991), avaliable in the `stats` package in base R, to select the bandwidth using pilot estimation of derivatives and calculate "peaks" in the density that correspond to stable rate values (Fig. 3c). The function then selects the rolling data windows that fall within the kernel density binwidth, merges all overlapping windows into one, and performs a linear regression on each new window.

>>>>

Metabolic rate is a fundamental trait linked to virtually all other biological functions, and is key to the prediction of patterns in ecology and conservation biology, from populations (Seibel and Drazen, 2007, Barneche et al., 2014) to ecosystems (Brown et al., 2004). It has been widely used to investigate the effects of external stressors on an organism, and has seen increased popularity in studies involving climate change where warming is expected to drive increased metabolism in ectotherms (Pörtner, 2002, McElroy et al., 2012, Carey et al., 2016, Delorme & Sewell, 2016). By far the most common metric used to measure physiological performance in animals, it can be applied from the whole organism to the level of cells and tissues (White & Kearney, 2013), and is typically quantified using respirometry, which measures oxygen uptake over time (Lighton, 2008).

There are at least four broad methodological approaches in respirometry: closed, intermittent flow, flowthrough and open. In closed respirometry, $O_2$ decrease is measured within a hermetically sealed chamber of known volume, sometimes set within a closed loop to allow circulation or mixing of the environment within the chamber. The oxygen recordings may be continuous through use of an oxygen probe, periodic through withdrawing water or gas samples at set intervals, or a two-point measurement consisting of the initial and final concentrations. Metabolic rates are estimated from respirometry by assuming a linear relationship between variables, and estimates of metabolic rate are straightforward in constant volume respirometry using the equation: $$RO_2 = \Delta O_2(V)$$ where $\Delta O_2$ is the slope of the regression that relates O~2~ concentration to time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time, and $V$ is the volume of the container (Lighton, 2008).

@@

Flowthrough respirometry involves a closed chamber, but with a regulated flow of air or water through it at a precisely determined rate. After equilibrium has been achieved, the oxygen concentration differential between the incurrent and excurrent channels, along with the flow rate, allows calculation of the oxygen extracted from the flow volume per unit time: $$VO_2 = (C_iO_2 - C_eO_2)FR$$ where $VO_2$ is the rate of $O_2$ consumption over time, $C_iO_2$ and $C_eO_2$ are the incurrent and excurrent $O_2$ concentrations, and $FR$ is the flow rate of water through the system  (Lighton, 2008).

A final method is open respirometry, in which an open tank or semi-enclosed area is used, but the input or mixing rate of oxygen from the surroundings is known or found to be negligible relative to oxygen consumption of the specimens (Leclercq et al., 1999). It is seldom used, but for some applications it is a sufficient and practical methodology (e.g. Gamble et al., 2014). The common equation used for open respirometry is: $$RO_2 = \Delta O_2(V) + \phi_d$$ where $\Delta O_2$ is the slope of the regression that relates $O_2$ concentration to time, $V$ is the volume of the container and $\phi_d$ is the air-sea oxygen flux as determined by Fick's Law (Leclercq et al., 1999).

@@

Data should be formatted correctly before use in `respR`. The function `check.input()` performs error checks on a data frame of any size, and can extract a two-column data frame for data exploration and subsequent analyses. Data must be numeric, because `respR` has limited support for date/time data (e.g. `POSIXct` and `POSIXlt` classes) and performing calculations on different time formats have many caveats that are beyond the scope of the package. Time data should be sequential, without duplicates, as such errors strongly indicate underlying sampling issues. Ideally, time data should also be evenly-spaced, that is, the time period between samples are equal. This has important ramifications on data extraction techniques later on, as substantial errors may occur in the event that calculations are based on the number of rows instead of time. A plot of the data is also shown for visual inspection (e.g. Fig. 1).

@@

Background $RO_2$ is measured to account for the influence of bacterial respiration in the chamber, and once measured, is often applied as a correction to the main experiment (Rodgers et al., 2016). Since background rates typically account for a small percentage of experimental rates, multiple “blank” experiments are routinely conducted alongside main experiments, and the rates are averaged across several datasets to obtain a more accurate estimate of the correction. The function `calc.bg.rate()` uses simple linear regression to simultaneously process multiple background rate measurements, as long as they share the time data, and produces an output object that can be used in later functions that accept background rate arguments.

@@

A typical procedure in the analysis respirometry data is to select one or more linear sections of the data for calculations of $RO_2$ (e.g. Gordon et al., 1989, Dorey et al., 2014, Chu & Gal, 2017). The function `calc.rate()` can manually extract one or more segments from a single data frame to process them simultaneously. Data segments can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a regular 10-minute time window across multiple specimens). Rates are determined using linear regression, however a simple delta approach is also included alongside all regression results, similar to a two-point analysis which is simply the difference in $O_2$ concentration divided by time.

In addition, we can take advantage of automated methods in `auto.rate()`, which implements rolling regression and kernel density estimation algorithms to detect patterns in the data. From the output we can automatically obtain the maximum or minimum rate of a specified sample size, extract regressions at non-overlapping intervals (Fig. 2a), or detect and rank linear sections of the data by sample size. Unlike manual methods, where data is, likely, arbitrary selected, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the length of the data (Fig. 2b). Thus, a total of $(m−n)+1$ number of overlapping regressions are fitted, and we obtain a new dataset of regressions as a function of time (rolling regression, Fig. 3), which is then ranked by size to obtain maximum and minimum values.

To estimate the most linear rate, further computations are performed after rolling regressions. Logically, linear sections of the data should reflect a consistently stable (i.e. flat) value in the rolling regression data. The function `auto.rate()` takes advantage of this relationship to identify consistent rates by performing a kernel density estimate of the data, implementing the methods of Sheather & Jones (1991) -- available in the `stats` package in R -- to derive the bandwidth used to construct the "peaks" in the density that correspond to stable rate values. The bandwidth is then used to select the range of overlapping regressions around the peak, and the single section of the data is resampled once using linear regression.

## Flowthrough respirometry

Calculating oxygen consumption using flowthrough respirometry requires different inputs of incurrent and excurrent oxygen concentration and a measure of the rate of flow through the respirometer. The function `calc.rate.ft`
